% ---------------------------------------------------------------------
% Podstawowe ustawienia i pakiety
% ---------------------------------------------------------------------
\RequirePackage[l2tabu, orthodox]{nag} % Wykrywa przestarzałe i niewłaściwe
% sposoby używania LaTeXa. Więcej jest w l2tabu English version.
\documentclass[a4paper,11pt]{article}
% \usepackage{babel}
% {rozmiar papieru, rozmiar fontu}[klasa dokumentu]
% w języku angielskim.
% \usepackage[utf8]{inputenc} % Włączenie kodowania UTF-8, co daje dostęp
% % do polskich znaków.
\usepackage{fontspec}
\setmainfont{Latin Modern Roman}

\usepackage{polyglossia}
\setdefaultlanguage{english}
\setotherlanguage{hebrew}


\newfontfamily{\hebrewfont}{FreeSerif}[
  Extension=.otf,
  Path=/usr/local/texlive/2022/texmf-dist/fonts/opentype/public/gnu-freefont/,
  UprightFont=*,
  ItalicFont=*Italic,
  BoldFont=*Bold,
  BoldItalicFont=*BoldItalic,
]





% ------------------------------
% Podstawowe pakiety (niezwiązane z ustawieniami języka)
% ------------------------------
\usepackage{microtype} % Twierdzi, że poprawi rozmiar odstępów w tekście.
\usepackage{graphicx} % Wprowadza bardzo potrzebne komendy do wstawiania
% grafiki.
\usepackage{verbatim} % Poprawia otoczenie VERBATIME.
\usepackage{textcomp} % Dodaje takie symbole jak stopnie Celsiusa,
% wprowadzane bezpośrednio w tekście.
\usepackage{vmargin} % Pozwala na prostą kontrolę rozmiaru marginesów,
% za pomocą komend poniżej. Rozmiar odstępów jest mierzony w calach.
% ------------------------------
% MARGINS
% ------------------------------
\setmarginsrb
{ 0.7in}  % left margin
{ 0.6in}  % top margin
{ 0.7in}  % right margin
{ 0.8in}  % bottom margin
{  20pt}  % head height
{0.25in}  % head sep
{   9pt}  % foot height
{ 0.3in}  % foot sep



% ------------------------------
% Często przydatne pakiety
% ------------------------------
\usepackage{csquotes} % Pozwala w prosty sposób wstawiać cytaty do tekstu.
\usepackage{xcolor} % Pozwala używać kolorowych czcionek (zapewne dużo
% więcej, ale ja nie potrafię nic o tym powiedzieć).



% ------------------------------
% Pakiety do tekstów z nauk przyrodniczych
% ------------------------------
\let\lll\undefined % Amsmath gryzie się z językiem pakietami do języka
% polskiego, bo oba definiują komendę \lll. Aby rozwiązać ten problem
% oddefiniowuję tę komendę, ale może tym samym pozbywam się dużego Ł.
\usepackage[intlimits]{amsmath} % Podstawowe wsparcie od American
% Mathematical Society (w skrócie AMS)
\usepackage{amsfonts, amssymb, amscd, amsthm} % Dalsze wsparcie od AMS
% \usepackage{siunitx} % Dla prostszego pisania jednostek fizycznych
\usepackage{upgreek} % Ładniejsze greckie litery
% Przykładowa składnia: pi = \uppi
\usepackage{slashed} % Pozwala w prosty sposób pisać slash Feynmana.
\usepackage{calrsfs} % Zmienia czcionkę kaligraficzną w \mathcal
% na ładniejszą. Może w innych miejscach robi to samo, ale o tym nic
% nie wiem.



% ##########
% Tworzenie otoczeń "Twierdzenie", "Definicja", "Lemat", etc.
\newtheorem{theorem}{Twierdzenie}  % Komenda wprowadzająca otoczenie
% „theorem” do pisania twierdzeń matematycznych
\newtheorem{definition}{Definicja}  % Analogicznie jak powyżej
\newtheorem{corollary}{Wniosek}



% ---------------------------------------
% Pakiety napisane przez użytkownika.
% Mają być w tym samym katalogu to ten plik .tex
% ---------------------------------------
% \usepackage{latexgeneralcommands}
% \usepackage{mathcommands}
% % \usepackage{calculuscommands}
% \newcommand{\conca}{\textrm{conca}}
% \usepackage{SchwartzBooksCommands}  % Pakiet napisany m.in. dla tego pliku.



% ---------------------------------------------------------------------
% Dodatkowe ustawienia dla języka polskiego
% ---------------------------------------------------------------------
\renewcommand{\thesection}{\arabic{section}.}
% Kropki po numerach rozdziału (polski zwyczaj topograficzny)
\renewcommand{\thesubsection}{\thesection\arabic{subsection}}
% Brak kropki po numerach podrozdziału



% ------------------------------
% Ustawienia różnych parametrów tekstu
% ------------------------------
\renewcommand{\arraystretch}{1.2} % Ustawienie szerokości odstępów między
% wierszami w tabelach.



% ------------------------------
% Pakiet "hyperref"
% Polecano by umieszczać go na końcu preambuły.
% ------------------------------
\usepackage{hyperref} % Pozwala tworzyć hiperlinki i zamienia odwołania
% do bibliografii na hiperlinki.










% ---------------------------------------------------------------------
% Tytuł i autor tekstu
\title{Basic information about languages in Sacred Scipture}


% \date{}
% ---------------------------------------------------------------------










% ####################################################################
\begin{document}
% ####################################################################





% ######################################
\maketitle % Tytuł całego tekstu
% ######################################





\textbf{Attention!} My English is very bad and no good.










% ######################################
\section{Languages} % Tytuł całego tekstu
% ######################################



Basic fact. We are using Latin alphabet, which was created by the
Romans\footnote{Name \textit{latin}, comes from the fact that region of the
  city of Rome was called Latium.}. Bible was written in mainly three
languages each one have its own alphabet, so they basically unreadable to
the most of us. These languages are Old Hebrew, Old Aramaic and one of the
version of Greek language known as Koine Greek or Alexandrian Greek. Bible
was written over the span that is usually count as 1000 years, so languages
changes during this time. But this is separate topic

Try to read John Milton now and he is only 400 or so years old.

Old Hebrew and Old Aramaic are similar in few ways. Both you read from
right-to-left and they alphabet consists of symbols only for consonants.
This is a bit more complicated when you encounter things like Hebrew letter
\begin{hebrew} א \end{hebrew}, but basically, there is no vowels in the
text, reader must to add it himself/herself.

Around 900 years after the time of Jesus, things called \textit{nekkudots}
that show which vowel is where, was added to the Hebrew script. But things
are far more complicated.

Basic Old Hebrew alphabet: \begin{hebrew} א \end{hebrew} (alef),
\begin{hebrew} ב‎ \end{hebrew} (bet), \begin{hebrew} ג‎ \end{hebrew} (gimel),
\begin{hebrew} ד‎ \end{hebrew} (dalet), \begin{hebrew} ה‎ \end{hebrew} (he),
\begin{hebrew} ו‎ \end{hebrew} (waw), \begin{hebrew} ז \end{hebrew} (zayin),
\begin{hebrew} ח \end{hebrew} (chet), \begin{hebrew} ט‎ \end{hebrew} (tet),
\begin{hebrew} י \end{hebrew} (yod), \begin{hebrew} כ‎, ך \end{hebrew} (kaf),
\begin{hebrew} ל‎ \end{hebrew} (lamed), \begin{hebrew} מ, ם \end{hebrew}
(mem), \begin{hebrew} נ, ן‎ \end{hebrew} (nun), \begin{hebrew} ס‎ \end{hebrew}
(samech), \begin{hebrew} ע \end{hebrew} (ayin),
\begin{hebrew} פ, ף‎ \end{hebrew} (pe), \begin{hebrew} צ, ץ‎ \end{hebrew}
\begin{hebrew} ק \end{hebrew} (qof), \begin{hebrew} ר \end{hebrew} (resh),
\begin{hebrew} ש \end{hebrew} (shin), \begin{hebrew} ת \end{hebrew} (tav).

Very crude way of reading Hebrew: for letters other than
\begin{hebrew} א, ע‎ \end{hebrew} read first consonant in the name. Again,
you basically must remember where and which vowels are in the world.

I won't show you Aramaic alphabet, because it was already hard to make my
computer accept Hebrew alphabet. You can find it e.g. on Wikipedia page
``Aramaic alphabet''.

Basic Greek alphabet: A, $\alpha$ (alpha), B, $\beta$ (beta), $\Gamma$, $\gamma$ (gamma),
$\Delta$, $\delta$ (delta), E, $\varepsilon$ (epsilon), Z, $\zeta$ (zeta), H, $\eta$ (eta), $\Theta$, $\theta$
(theta), I, $\iota$ (iota), K, $\kappa$ (kappa), $\Lambda$, $\lambda$ (lambda), M, $\mu$ (mu), N,
$\nu$ (nu), $\Xi$, $\xi$ (xi), O, o (omicron), $\Pi$, $\pi$ (pi), $\Sigma$, $\sigma$, $\xi$
(sigma), T, $\tau$ (tau), Y, $\upsilon$ (upsilon), $\Phi$, $\varphi$ (phi), X, $\chi$, $\Psi$, $\psi$
(psi), $\Omega$, $\omega$ (omega).

Greek and Latin are languages two languages from which most of European
languages were derived, so they are ``as we expected''. You have vowels and
similar construction, but Greek lacks interconnection. Polish and English
were created as imitation of Latin in writing. Crude way of reading Greek:
just read first letter of the name.

There is much more to Hebrew and Greek that this letters. You can various
additional symbols, but since using them is beyond me, I will omit them.










% ######################################
\section{What all this names means?} % Tytuł całego tekstu
% ######################################



``Bible'' comes from Koine Greek $\tau \alpha$ $\beta \iota \beta \lambda \iota \alpha$, ``ta biblia'', meaning
``the books''\footnote{Note that in English we still have
  ``bibliography''.}. ``Sacred Scripture'' comes from Latin, it means
``sacred writing''\footnote{Note in English we have words as ``scribe'' and
  ``scribbling''.}.

Old Testament is called by Jews \begin{hebrew} תנך \end{hebrew} (Tanakh) or
\begin{hebrew} מקרא \end{hebrew} (Mikra). It is divided in three parts:
\begin{hebrew} תורה \end{hebrew} (Torah, en. ``Teachings''),
\begin{hebrew} נביאים \end{hebrew} (Nevi'im, en. ``Spokeperson'') and
\begin{hebrew} כתובים \end{hebrew} (Ketuvim, en. ``Writing'').

Torah \begin{hebrew} תורה \end{hebrew} contains Five Books of Moses. Nevi'im
\begin{hebrew} נביאים \end{hebrew} you can say it contains
prophets, but live is not so simple. I don't even try to characterize
Ketuvim \begin{hebrew} כתובים \end{hebrew}. Name ``Tanakh''
\begin{hebrew} תנך \end{hebrew} is quasi anagram of first letters of Tora,
Nevi'im and Ketuvim.

Five Books of Moses derives they Hebrew names from their first word.
\begin{itemize}

\item First. \begin{hebrew} בראשית \end{hebrew}, Bereshit, en. ``In the beginning''.
  In Greek $\Gamma \varepsilon \nu \varepsilon \sigma \iota \zeta$, Genesis, en. ``Creation''\footnote{Note words
    like ``genetics'' or ``genealogy''}.

\item Second. \begin{hebrew} שמות \end{hebrew}, Shemot, en. ``Names''. In Greek
  E$\xi$o$\delta$o$\zeta$, Exodos, en. ``Exit''.

\item Third. \begin{hebrew} וקרא \end{hebrew}, Vayikra, en. ``And He called''. In
  Greek $\Lambda \varepsilon \upsilon \iota \tau \iota \kappa$o$\nu$, Leutikon, en. ``Relating to the Levites''.

\item Fourth. \begin{hebrew} במדבר \end{hebrew}, Bemidbar, en ``In the desert
  [of]''. In Greek A$\rho \iota \theta \mu$o$\iota$, Arithmoi, en. ``Numbers''\footnote{Note
    words like ``arithmetics''.}

\item Fifth. \begin{hebrew} דברים \end{hebrew}, Devarim, en. ``Things''. In Greek
  $\Delta \varepsilon \upsilon \tau \varepsilon \rho$o$\nu$o$\nu$o$\mu \iota$o$\nu$, Deuteronomion, en. ``Second-law''.

\end{itemize}

It is worth of notice how English Standard Version translate the beginning
of the each book of Torah.

\begin{itemize}

\item Genesis, \textit{In the beginning, God created the heavens and the
    earth.}

\item Exodus, \textit{These are the names of the sons of Israel who came to
    Egypt with Jacob, each with his household: Reuben, Simeon, Levi, and
    Judah, Issachar, Zebulun, and Benjamin, Dan and Naphtali, Gad and
    Asher.}

\item Leviticus, \textit{The Lord called Moses and spoke to him from the
    tent of meeting, saying, “Speak to the people of Israel and say to
    them, When any one of you brings an offering to the Lord, you shall
    bring your offering of livestock from the herd or from the flock.}

\item Numbers, \textit{The Lord spoke to Moses in the wilderness of Sinai,
    in the tent of meeting, on the first day of the second month, in the
    second year after they had come out of the land of Egypt, saying,
    “Take a census of all the congregation of the people of Israel, by
    clans, by fathers’ houses, according to the number of names, every
    male, head by head.}

\item Deuteronomy, \textit{These are the words that Moses spoke to all
    Israel beyond the Jordan in the wilderness, in athe Arabah opposite
    Suph, between Paran and Tophel, Laban, Hazeroth, and Dizahab.}

\end{itemize}

Of course live and Hebrew cannot be so simple. We are only scratching the
surface of it. For example name of the first book of the Bible is
\begin{hebrew} בראשית \end{hebrew} Bereshit, which we translate ``In the
beginning'', but literally means ``In the head''. It is correct, since ``In
the head'' in Hebrew also mean ``starting point''. This language is full of
such constructions.










% ############################
\subsection{Two kinds of Torah} % Tytuł całego tekstu
% ############################



Among Jews another name of Five Books of Moses is \begin{hebrew} תורה שבבתב
\end{hebrew}, Torah Shebikhtav, en. ``written instructions'' to distinguish
from \begin{hebrew} תורה שבעל פה \end{hebrew}, Torah Shebeal Peh, ``oral
instructions''.

I once find opinion about conundrum from Gospel of Matthew. It seems that
some quotations of the prophets in this Gospel cannot be localized in Old
Testament. At least I don't know any convincing identification of them. So
explanation given is that Matthew as typical Jews of his time, considered
only part of the ``instructions'' and ``prophecies'' being written down in
Old Testament, other are preserve only in oral form, but it doesn't mean to
that they are not part of divine revelation.

This is against doctrine of ``Sola Scriptura'', but not mentioning it would
be very bad thing on my side. And I need to mention that ``Sola Scriptura''
is a~phrase in Latin.






\begin{hebrew} א ב‎ \end{hebrew} (bet), \begin{hebrew} ג‎ \end{hebrew} (gimel),
\begin{hebrew} ד‎ \end{hebrew} (dalet), \begin{hebrew} ה‎ \end{hebrew} (he),
\begin{hebrew} ו‎ \end{hebrew} (waw), \begin{hebrew} ז \end{hebrew} (zayin),
\begin{hebrew} ח \end{hebrew} (chet), \begin{hebrew} ט‎ \end{hebrew} (tet),
\begin{hebrew} י \end{hebrew} (yod), \begin{hebrew} כ‎, ך \end{hebrew} (kaf),
\begin{hebrew} ל‎ \end{hebrew} (lamed), \begin{hebrew} מ, ם \end{hebrew},
\begin{hebrew} נ, ן‎ \end{hebrew} (nun), \begin{hebrew} ס‎ \end{hebrew}
(samech), \begin{hebrew} ע \end{hebrew} (ayin),
\begin{hebrew} פ‎, ף‎ \end{hebrew} (pe), \begin{hebrew} צ‎, ץ \end{hebrew}
(tsadi), \begin{hebrew} ק \end{hebrew} (Qof), \begin{hebrew} ר \end{hebrew}
(resh), \begin{hebrew} ש \end{hebrew} (shin), \begin{hebrew} ת \end{hebrew}
(tav).










% ############################
\subsection{New Testament} % Tytuł całego tekstu
% ############################



New Testament is to us know only in Koine Greek, so outside translation
coming from XVI century onward, there is no Hebrew names for the books of
New Testament.

New Testament in Greek is H K$\alpha \iota \nu \eta$ $\Delta \iota \alpha \theta \eta \kappa \eta$, He Kaine Diatheke.
Word ``diatheke'' main meaning is posthumous will, so ``New Testament'' is
good translation, at least in polish\footnote{``Testametn'', of course, is
  world taken from Latin.}.

Origin of this use of $\Delta \iota \alpha \theta \eta \kappa \eta$ comes back to the most important
translation of Hebrew books of Old Testament into Greek. I happens somewhere
between 400 BC and 100 BC and it is know as Septuaginta\footnote{From Latin
  it means ``seventy'', so also know as 70 and LXX.}.

When in Genesis Abram make covenant with God, words used for ``covenant'' is
\begin{hebrew} ברית \end{hebrew}, brit. In Septuaginta this word is
translated ad $\Delta \iota \alpha \theta \eta \kappa \eta$, diatheke. Don't ask me why.

The most probably version of origin of this name for 27 books, that I know,
is as follow. When Christ broke the bread saying ``This is my body'' in
Greek version of Gospels He call this ceremony (?)
H K$\alpha \iota \nu \eta$ $\Delta \iota \alpha \theta \eta \kappa \eta$, He Kaine Diatheke, New Testament, New
Covenant. After that this name was associated with books that were read
during ceremony of the New Testament, so they were also start being called
New Testament.










% ############################
\subsection{Gospels} % Tytuł całego tekstu
% ############################



It seems that ``Gospel'' is a name that originates in English. In Greek we
have E$\upsilon \alpha \gamma \gamma \varepsilon \lambda \iota$0$\nu$, Euangelion, where ``E$\upsilon$'', ``Eu'' means ``good''
and ``$\alpha \gamma \gamma \varepsilon \lambda$o$\zeta$'', ``angelion'' means ``message'', literally ``good
message''. In polish it is ``Ewangelia'' in German ``Evangelium''.








% ######################################
\section{Songs} % Tytuł całego tekstu
% ######################################



Depending on whom published your Bible,




% #####################################################################
% #####################################################################
% Bibliografia
\bibliographystyle{plalpha}

\bibliography{MathComScienceBooks}{}





% ############################

% Koniec dokumentu
\end{document}
